---
title: "CAI Negotiation Assignment - Report on a combination of runs"
author: "Group 33"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    keep_tex: yes
    number_sections: yes
header-includes:
  - \usepackage{times}
  - \usepackage{ijcai25}
---

\tableofcontents

# About

This document is a report on a single combination of agents and domain.
The task is run twice for domain symmetry.

```{r setup, echo=FALSE}
source("aana_utils.R")
```

```{r, echo=FALSE, results='asis'}
repo_path <- "../cai-anl-33/"
venv_dir <- "venv"

# !!
# CHANGE THIS to simulate again. (or delete the results files)
# !!
run_label <- "run02"

# slashes are important :o
run_dir <- "single_runs/"

agents = c(
  "agents.ANL2022.dreamteam109_agent.dreamteam109_agent.DreamTeam109Agent",
  "agents.ANL2022.micro_agent.micro_agent.micro_agent.MiCROAgent"
)
domain = "domain01"

cat(sprintf("\nThe run label is \\verb|%s|.", run_label))
```

```{r run_agent_left, results='hide', echo=FALSE}

run_name <- paste(run_label, "left", sep='-')

if (!dir.exists(paste(run_dir, run_name, sep=''))) {
  cat("\nRunning agents: \n", file=stderr())
  cat(paste(agents, sep="\n"), file=stderr())
  cat("\nThis shouldn't take more than 10 seconds...\n", file=stderr())
  run_agents(repo_path, "venv", "single_runs/", run_name, 
      agents[1],
      agents[2],
      paste("domains", domain ,"profileA.json", sep='/'),
      paste("domains", domain ,"profileB.json", sep='/')
  )
} else {
  cat("\nNot running, results exist. To re-simulate, delete them or change the run name.\n", file=stderr())
}

```

```{r, results="asis", echo=FALSE, warning=FALSE, fig.height=5.1, fig.width=6, fig.align='center'}

cat("\n# Results for the left-hand-side run\n")


left_data <- try(
  read_results(run_dir, run_name)
)

if (inherits(left_data, "try-error")) {
  cat(geterrmessage())
} else {
  plot_run_results(left_data)
}
```

```{r run_agent_right, results='hide', echo=FALSE}

run_name <- paste(run_label, "right", sep='-')

if (!dir.exists(paste(run_dir, run_name, sep=''))) {
  cat("\nRunning agents: \n", file=stderr())
  cat(paste(agents, sep="\n"), file=stderr())
  cat("\nThis shouldn't take more than 10 seconds...\n", file=stderr())
  run_agents(repo_path, "venv", "single_runs/", run_name, 
      agents[1],
      agents[2],
      paste("domains", domain ,"profileB.json", sep='/'),
      paste("domains", domain ,"profileA.json", sep='/')
  )
} else {
  cat("\nNot running, results exist. To re-simulate, delete them or change the run name.\n", file=stderr())
}

```

```{r, results="asis", echo=FALSE, warning=FALSE, fig.height=5.1, fig.width=6, fig.align='center'}

cat("\n# Results for the right-hand-side run\n")

right_data <- try(
  read_results(run_dir, run_name)
)

if (inherits(right_data, "try-error")) {
  cat(geterrmessage())
} else {
  plot_run_results(right_data)
}
```

# Evaluation

After running the two agents on the same domain from both ends, we can identify
(by taking the utility difference) which agent seems to have an advantage.

```{r, results="asis", echo=FALSE}

if (inherits(left_data, "try-error")) {
  final1_left <- 0
  final2_left <- 0
} else {
  util1_left <- left_data[[1]]$Utility1
  util2_left <- left_data[[1]]$Utility2
  n_left <- length(util1_left)
  final1_left <- util1_left[n_left]
  final2_left <- util2_left[n_left]
}

# flipped because of switched agent positions
if (inherits(right_data, "try-error")) {
  final1_right <- 0
  final2_right <- 0
} else {
  util1_right <- right_data[[1]]$Utility1
  util2_right <- right_data[[1]]$Utility2
  n_right <- length(util1_right)
  final1_right <- util1_right[n_right]
  final2_right <- util2_right[n_right]
}

total_util1 = final1_left + final1_right
total_util2 = final2_left + final2_right

agent_labels = sapply(agents, agent_name)

```

Final utility values:
\begin{itemize}
  \item `r agent_labels[1]`: `r total_util1`
  \item `r agent_labels[2]`: `r total_util2`
\end{itemize}